---
title: "Pre Analysis Plan: Chakraborty (2021) Reproduction Study"
author: "Samuel Barnard"
date: "`r Sys.Date()`"
output: html_document
editor_options:
  markdown:
    wrap: sentence
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../../docs") })
nocite: '@*'
---

Version 0.1 \| First Created March 10, 2025 \| Updated March 10, 2025

# Abstract

Chakraborty (2021) investigates the relationships between COVID-19 incidence rates and several demographic characteristics of people with disabilities by county in the continental United States. The aim of the study is to investigate whether people with disabilities (PwDs) face disproportionate challenges due to COVID-19.

Continued, from **"Reproduction of Chakraborty 2021: An intracategorical analysis of COVID-19 and people with disabilities"** (Holler et al.):
To do so, Chakraborty examines the statistical relationship between county incidence rates of COVID-19 cases and county-level percentages of people with disabilities and different socio-demographic characteristics.
Specifically, Chakraborty tests county-level bivariate correlations between COVID-19 incidence against the percentage of disability as one hypothesis, and tests correlation between COVID-19 incidence and percentage of people with disabilities in 18 different socio-demographic categories of race, ethnicity, poverty status, age, and biological sex.
Chakraborty then re-tests for the same county-level associations while controlling for spatial dependence.
Spatial dependence is controlled by constructing generalized estimating equation (GEE) models using a combination of state and spatial clusters of COVID-19 incidence as to define the GEE clusters.
One GEE model is constructed for each of the four types of socio-demographic category: race, ethnicity, age, and biological sex.
Chakraborty (2021) finds significant positive relationships between COVID-19 rates and socially vulnerable demographic categories of race, ethnicity, poverty status, age, and biological sex.

# Study Design

This study is a reproduction, with the goal of examining Chakraborty's study design and its impact particularly in public policy as well as in fields such as research and education. This reproduction will attempt to reproduce the original study's results.

This will include the map of county level distribution of COVID-19 incidence rates (Fig. 1), the summary statistics for disability and socio-demographic variables and bivariate correlations with county-level COVID-19 incidence rate (Table 1), and the GEE models for predicting COVID-19 county-level incidence rate (Table 2).
A successful reproduction should be able to generate identical results as published by Chakraborty (2021).

Chakraborty, J.
2021.
Social inequities in the distribution of COVID-19: An intra-categorical analysis of people with disabilities in the U.S.
*Disability and Health Journal* 14:1-5.
<https://doi.org/10.1016/j.dhjo.2020.101007>

# Study data

## [Study metadata](data/metadata/)

## ACS Socio-demographic data

The American Community Survey (ACS) five-year estimate (2014-2018) variables used in the study are outlined in the table below.
Details on ACS data collection can be found at <https://www.census.gov/topics/health/disability/guidance/data-collection-acs.html> and details on sampling methods and accuracy can be found at <https://www.census.gov/programs-surveys/acs/technical-documentation/code-lists.html>.

| Variable Name in Study | ACS Variable name |
|:----------------------------------:|:----------------------------------:|
| percent of total civilian non-institutionalized population with a disability | S1810_C03_001E |
| **Race** |  |
| percent w disability: White alone | S1810_C03_004E |
| percent w disability: Black alone | S1810_C03_005E |
| percent w disability: Native American | S1810_C03_006E |
| percent w disability: Asian alone | S1810_C03_007E |
| percent w disability: Other race | S1810_C03_009E |
| **Ethnicity** |  |
| percent w disability: Non-Hispanic White | S1810_C03_0011E |
| percent w disability: Hispanic | S1810_C03_012E |
| percent w disability: Non-Hispanic non-White | (S1810_C02_001E - S1810_C02_011E - S1810_C02_012E) / (S1810_C01_001E - S1810_C01_011E - S1810_C01_012E) \* 100 |
| percent w disability: Other race | S1810_C03_009E |
| **Poverty** |  |
| percent w disability: Below poverty level | (C18130_004E + C18130_011E + C18130_018E) / C18130_001E \* 100 |
| percent w disability: Above poverty level | (C18130_005E + C18130_012E + C18130_019E) / C18130_001E \* 100 |
| **Age** |  |
| percent w disability: 5-17 | S1810_C03_014E |
| percent w disability: 18-34 | S1810_C03_015E |
| percent w disability: 35-64 | S1810_C03_016E |
| percent w disability: 65-74 | S1810_C03_017E |
| percent w disability: 75+ | S1810_C03_018E |
| **Biological sex** |  |
| percent w disability: male | S1810_C03_001E |
| percent w disability: female | S1810_C03_003E |

: Disability Subgroup Variables

American Community Survey (ACS) data for sociodemographic subcategories of people with disabilities can be accessed by using the `tidycensus` package to query the Census API. This requires an API key which can be acquired at [api.census.gov/data/key_signup.html](https://api.census.gov/data/key_signup.html).

## COVID-19 data

Data on COVID-19 cases from the Johns Hopkins University dashboard have been provided directly with the research compendium because the data is no longer available online in the state in which it was downloaded on August 1, 2020.
The dashboard and cumulative counts of COVID-19 cases and deaths were continually updated, so an exact reproduction required communication with the original author, Jayajit Chakraborty, for assistance with provision of data from August 1, 2020.
The data includes an estimate of the total population (`POP_ESTIMA`) and confirmed COVID-19 cases (`Confirmed`).
The COVID-19 case data expresses cumulative count of reported COVID-19 from 1/22/2020 to 8/1/2020.
Although metadata for this particular resource is no longer available from the original source, one can reasonably assume that the total population estimate was based on the 2014-2018 5-year ACS estimate, as the 2019 estimates data had not been released yet.

Versions of the data can be found at the John Hopkins CCSE COVID-19 Data Repository (<https://github.com/CSSEGISandData/COVID-19>).
However, archived data only provides summaries at the national scale.
We received the COVID-19 case data through 8/1/2020 at the county level from the author, as there is no readily apparent way to access archived data from the Johns Hopkins University Center for Systems Science Engineering database.

# Materials and procedure

## Computational environment
```{r setup, include = FALSE}
# list of required packages
packages <- c(
  "tidycensus", "tidyverse", "downloader", "sf", "classInt", "readr",
  "here", "s2", "pastecs", "tmap", "SpatialEpi", "svDialogs",
  "geepack", "knitr", "kableExtra", "foreign", "broom", "dotwhisker", "dotenv", "tmap"
)

# load and install required packages
if(!require(groundhog)){
  install.packages("groundhog")
  require(groundhog)
}

if(!require(here)){
  install.packages("here")
  require(here)
}

groundhog.day <- "2025-02-01"
set.groundhog.folder(here("data", "scratch", "groundhog"))

groundhog.library(packages, groundhog.day,
                  tolerate.R.version='4.4.2')
# you may need to...
# install a correct version of R
# install the rstudioapi package with install.packages("rstudioapi")
# respond OK in the console to permit groundhog to install packages
# restart the R session and rerun this code to load installed packages
# In RStudio, restart r with Session -> Restart Session

# non-groundhog method for installing packages:
# lapply(packages, library, character.only = TRUE)

# save the R processing environment
writeLines(
  capture.output(sessionInfo()),
  here("procedure", "environment", paste0("r-environment-", Sys.Date(), ".txt"))
)

# set up default knitr parameters
knitr::opts_chunk$set(
  echo = FALSE,
  fig.width = 8,
  fig.path = paste0(here("results", "figures"), "/")
)
```

## Prior observations

As of the release of this pre-analysis plan, I have not observed the data past recording the metadata above.
My experience with the methods of this study stem from an in-class analysis and breakdown of the original study to better understand data sources and transformations.
In-class discussion included exposure to some workflow from a previous reproduction study ("Reproduction of Chakraborty 2021: An intracategorical analysis of COVID-19 and people with disabilities"), which among other things helped to solidify important steps in data acquisition and transformation, for instance the correct data tables from the ACS to query. This prior reproduction was also completed by contacting the original study's researchers for clarification of their process. 

## Data transformations and Analysis

### Step 1: Acquire John's Hopkins COVID data

Examine ACS variables from relevant tables

```{r}
acs_subject_vars <- load_variables(2018, "acs5/subject")
acs_vars_S1810 <- acs_subject_vars |> dplyr::filter(str_detect(name, "S1810"))
write_csv(acs_vars_S1810, here("data", "metadata", "acs_S1810_vars.csv"))

acs_vars <- load_variables(2018, "acs5")
acs_vars_C18130 <- acs_vars |> dplyr::filter(str_detect(name, "C18130"))
write_csv(acs_vars_C18130, here("data", "metadata", "acs_C18130_vars.csv"))
```

Query ACS: data at the county level, for the continental US only.

```{r API-Load-ACS, eval=FALSE}
# Query disability demographic data with geographic boundaries
acs <- get_acs(
  geography = "county",
  table = "S1810",
  year = 2018,
  output = "wide",
  cache_table = TRUE,
  geometry = TRUE,
  keep_geo_vars = TRUE
)

# Query poverty and disability data
acs_pov <- get_acs(
  geography = "county",
  table = "C18130",
  year = 2018,
  output = "wide",
  cache_table = TRUE
)

# Query state geographic data
state <- get_acs(
  geography = "state",
  year = 2018,
  variables = c("B01001_001"),
  geometry = TRUE,
  keep_geo_vars = TRUE
)

# Save query results
saveRDS(acs, here("data", "raw", "public", "acs.RDS"))
saveRDS(acs_pov, here("data", "raw", "public", "acs_pov.RDS"))
saveRDS(state, here("data", "raw", "public", "state.RDS"))
```

Load ACS data

```{r load-acs}
acs <- readRDS(here("data", "raw", "public", "acs.RDS"))
acs_pov <- readRDS(here("data", "raw", "public", "acs_pov.RDS"))
state <- readRDS(here("data", "raw", "public", "state.RDS"))
```

Transformations: to be consistent with study extent, exclude Alaska, Hawai'i and Puerto Rico from the dataset.

Also, join poverty ACS variable to other ACS data using GEOID to create one dataset with all of the study variables.

Finally, transform the ACS geographic data into Contiguous USA Albers Equal Area projection and fix geometry errors.

```{r filter-join-acs}
# Remove Alaska, Hawaii & Puerto Rico,
# transform coordinate system and fix geometries
acs <- dplyr::filter(acs, !STATEFP %in% c("02", "15", "72")) %>%
  st_transform(5070) %>%
  st_make_valid()

# Remove Alaska, Hawaii & Puerto Rico,
state <- dplyr::filter(state, !STATEFP %in% c("02", "15", "72")) %>%
  st_transform(5070)

# Join poverty data to disability data
acs <- left_join(acs, acs_pov, by = "GEOID")
rm(acs_pov)
```

### Step 2: Calculate incident rate by county

Load COVID data

```{r load-covid-data}
covid <- read_sf(here("data", "raw", "public", "covidcase080120.gpkg"))

# select and rename the fips code, population, cases, and x,y coordinates
covid <- select(covid,
  fips = FIPS,
  pop = POP_ESTIMA,
  cases = Confirmed,
  x = X, y = Y
)
```

Calculate incidence rate (cases per 100,000 people)

Convert the COVID data to a non-geographic data frame.

```{r covid-rate}
covid_table <- covid %>%
  mutate(covid_rate = round(covid$cases / covid$pop * 100000, 2)) %>%
  st_drop_geometry()
```

### Step 3: Acquire ACS data

Join variables from `acs_vars_S1810` and `acs5_c18130`

### Step 4: Join ACS to COVID data

Join dependent COVID data to independent ACS demographic data.

```{r join-covid-to-acs}
# Join COVID incidence rate data to acs data
acs_covid <- acs_derived %>%
  left_join(covid_table, by = "fips")

# move covid_rate column prior to disability percentages
acs_covid <- acs_covid %>%
  select(fips, statefp, county, county_st, covid_rate, everything())

rm(acs, acs_derived, covid)
```

**Unplanned deviation for reproduction**: There is one county with missing disability and poverty data.
This was not mentioned in the original study or in our pre-analyis plan.
However, we replace the missing data with zeros, producing results identical to Chakraborty's.

```{r missing data}
# county with missing data
dplyr::filter(acs_covid, is.na(bpov_pct)) %>% st_drop_geometry() %>% kable()

# replace NA with 0 for missing data
acs_covid[is.na(acs_covid$bpov_pct), ]$bpov_pct <- 0
acs_covid[is.na(acs_covid$apov_pct), ]$apov_pct <- 0
```

### Map: COVID incident rate data

```{r map-covid-rates, message = FALSE}
tmap_mode(mode = "plot")

tm_covid_rates <- tm_shape(acs_covid) +
  tm_polygons("covid_rate",
    title = "COVID-19 Cases per 100,000 people\n(22 January 2020 to 1 August 2020)",
    style = "quantile",
    border.alpha = .2,
    lwd = 0.2,
    palette = "Oranges",
  ) +
  tm_shape(state) +
    tm_borders("grey", lwd = .2) +
  tm_layout(
    legend.position = c("left", "bottom"),
    legend.title.size = 0.5,
    legend.text.size = 0.5
  )

tm_covid_rates #render map
```

### Step 5: Mutate PwD variables

Calculate independent socio-demographic variables of people with disabilities as percentages for each sub-category of disability (race, ethnicity, poverty, age, and biological sex) and remove raw census data from the data frame (workflow step 4).
Reproject the data into an Albers equal area conic projection.

Number of PwDs / total from each PwD variable * 100

```{r Preprocess-ACS}
# calculate percentages
acs_derived <- mutate(acs,
  dis_pct = S1810_C02_001E / S1810_C01_001E * 100,
  white_pct = S1810_C02_004E / S1810_C01_001E * 100,
  black_pct = S1810_C02_005E / S1810_C01_001E * 100,
  native_pct = S1810_C02_006E / S1810_C01_001E * 100,
  asian_pct = S1810_C02_007E / S1810_C01_001E * 100,
  other_pct =
    (S1810_C02_008E + S1810_C02_009E + S1810_C02_010E) / S1810_C01_001E * 100,
  non_hisp_white_pct = S1810_C02_011E / S1810_C01_001E * 100,
  hisp_pct = S1810_C02_012E / S1810_C01_001E * 100,
  non_hisp_non_white_pct =
    (S1810_C02_001E - S1810_C02_012E - S1810_C02_011E) / S1810_C01_001E * 100,
  bpov_pct = (C18130_004E + C18130_011E + C18130_018E) / C18130_001E * 100,
  apov_pct = (C18130_005E + C18130_012E + C18130_019E) / C18130_001E * 100,
  pct_5_17 = S1810_C02_014E / S1810_C01_001E * 100,
  pct_18_34 = S1810_C02_015E / S1810_C01_001E * 100,
  pct_35_64 = S1810_C02_016E / S1810_C01_001E * 100,
  pct_65_74 = S1810_C02_017E / S1810_C01_001E * 100,
  pct_75 = S1810_C02_018E / S1810_C01_001E * 100,
  male_pct = S1810_C02_002E / S1810_C01_001E * 100,
  female_pct = S1810_C02_003E / S1810_C01_001E * 100
)

# select only relevant geographic identifiers and derived percentages
acs_derived <- acs_derived %>%
  select(
    fips = GEOID,
    statefp = STATEFP,
    county = NAME.x,
    county_st = NAME,
    contains("pct")
  )
```

### Map: PwD rates by county

**Unplanned deviation for reproduction**: We also map the spatial distribution of the percent of people with any disability to improve our understanding of the geographic patterns and relationships of between the overarching independent variable (percentage of people with disability) and the dependent variable (COVID-19 incidence rate).

```{r map-disability-rates, message = FALSE}
tm_disability_rates <- tm_shape(acs_covid) +
  tm_polygons("dis_pct",
    title = "Percent of People with Disability\n(ACS 2014-2018)",
    style = "quantile",
    border.alpha = .2,
    lwd = 0.2,
    palette = "Blues"
  ) +
  tm_shape(state) +
  tm_borders("grey", lwd = .5) +
  tm_layout(
    legend.position = c("left", "bottom"),
    legend.title.size = 0.5,
    legend.text.size = 0.5
  )

tm_disability_rates
```

### Step 6: `summarise()`

Calculate summary statistics for dependent COVID-19 rate and independent PwD socio-demographic characteristics (Min, Max, Mean, SD) to reproduce columns of original study table 1

```{r descriptive-statistics}
acs_covid_stats <- acs_covid %>%
  st_drop_geometry() %>%
  select(covid_rate, contains("pct")) %>%
  stat.desc(norm = TRUE) %>%
  round(2) %>%
  t() %>%
  as.data.frame() %>%
  select(min, max, mean, SD = std.dev)

acs_covid_stats %>%
  kable(caption = "Reproduced Descriptive Statistics",
        align = "c")
```

### Step 7: Bivariate Pearson product-moment correlations

Correlate `cor()` COVID cases, pct disability for each disability variable

#### `Mutate()` to produce Pearson's r column in Table 1

`mutate( t = abs(r) / sqrt((1 - r\^2) / (df)), p = pt(t, df, lower.tail = FALSE) )`

```{r pearsons-correlation}
df <- sum(!is.na(acs_covid$dis_pct)) - 2

pearsons_r <- acs_covid %>%
  select(where(is.numeric)) %>%
  st_drop_geometry() %>%
  cor(method = "pearson", use = "everything") %>%
  as.data.frame() %>%
  select(r = covid_rate) %>%
  mutate(
    t = abs(r) / sqrt((1 - r^2) / (df)),
    p = pt(t, df, lower.tail = FALSE)
  ) %>%
  round(3) %>%
  rownames_to_column("variable") %>%
  dplyr::filter(variable != "covid_rate")

pearsons_r %>%
  kable(caption = "Reproduced Pearson's R",
        align = "c") %>%
  column_spec(2:4, width = "4em") %>%
  kable_styling(full_width = FALSE)
```

### Check results with Table 1

#### Digitize Table 1 from original study

Subtract results from Chakraborty's Table 1

Compare reproduced descriptive statistics to original descriptive statistics.
Difference is calculated as 'reproduction study - original study'.
Identical results will result in zero.

```{r compare-descriptive-stats}
# load original table 1 results
table1 <- read.csv(here("data", "raw", "public", "chakraborty", "table1.csv"))

# subtract original results from reproduced results
(select(acs_covid_stats, min, max, mean, SD) -
  select(table1, min, max, mean, SD)) %>%
  kable(caption = "Descriptive Statistics Comparison",
        align = "c") %>%
  column_spec(2:5, width = "4em") %>%
  kable_styling(full_width = FALSE)

rm(acs_covid_stats)
```

Compare the reproduced Pearson's *r* correlation coefficients to the original study's Pearson's *r* correlation coefficients.
Stars indicates the significance level with two stars for `p < 0.01` and one star for `p < 0.05`.
Correlation difference `rp_r_diff` is calculated between the reproduction study `rp_r` and original study `or_r` as `rp_r_diff = rp_r - or_r` Direction difference `rp_dir_diff` is calculated as `(rp_r > 0) - (or_r > 0)`, giving `0` if both coefficients have the same direction, `1` if the reproduction is positive and the original is negative, and `-1` if the reproduction is negative but the original is positive.

```{r compare-pearsons-correlation}
# calculate number of significance stars at p < 0.01 and p < 0.05 levels.
pearsons_r <- mutate(pearsons_r, rp_stars = as.numeric(as.character(cut(p,
  breaks = c(-0.1, 0.01, 0.05, 1),
  labels = c(2, 1, 0)
))))

# join reproduction coefficients to original study coefficients
correlations <- table1 %>%
  dplyr::filter(variable != "covid_rate") %>%
  select(variable, or_r = r, or_stars = stars) %>%
  left_join(select(pearsons_r, variable, rp_r = r, rp_stars), by = "variable")

# find difference between coefficient and stars
correlations <- correlations %>%
  bind_cols(rename_with(
    correlations[, 4:5] - correlations[, 2:3],
    ~ paste0(.x, "_diff")
  ))

# find coefficients with different directions
correlations <- correlations %>% mutate(rp_dir_diff = (rp_r > 0) - (or_r > 0))

correlations %>%
  kable(caption = "Compare reproduced and original Pearson's R",
        col.names = c("Variable", "R", "Sig. Level", "R", "Sig. Level", "R", "Sig. Level", "Direction"),
        align = "c") %>%
  kable_styling() %>%
  add_header_above(c(" " = 1, "Original" = 2, "Reproduced" = 2, "Difference" = 3))
```

Reproduction correlation coefficients varied slightly from the original study coefficients by +/- 0.006.
All but one Pearson's correlation coefficient was significant to the same level, and the exception was age 18 to 34.
Counter-intuitively, the correlation coefficient was slightly closer to 0 but the *p* value was also found to be more significant, suggesting a difference in the estimation of *t* and/or *p*, or a typographical error.
All of the coefficients had the same direction.

## *Planned Deviation*

### Step 8: Kulldorff method with `SpatialEpi` package

First, calculate the Kulldorff spatial scan statistic using SpatialEpi.
Optionally, skip this code block due to long run times of more than 10 minutes.

```{r SpatialEpi-Kulldorff, eval = FALSE, fig.width=4, fig.height=4}
start_time <- Sys.time()
covid_geo <- covid_table %>%
  select(x, y) %>%
  latlong2grid()
# latlong2grid creates approximate equidistant cylindrical grid
# could probably reproject to epsg 5070 and create table with x and y

# calculate expected cases with one strata
expected.cases <- expected(covid_table$pop, covid_table$cases, 1)

# Kulldorff spatial scan statistic
covid_kulldorff <- kulldorff(
  geo = covid_geo,
  cases = covid_table$cases,
  population = covid_table$pop,
  expected.cases = expected.cases,
  pop.upper.bound = 0.5,
  n.simulations = 999,
  alpha.level = 0.05,
  plot = TRUE
)

print(
  paste(
    "Run time:",
    round(difftime(Sys.time(), start_time, units = "mins"), 2),
    "minutes"
  ),
  quote = FALSE
)
rm(covid_geo, expected.cases, start_time)

# save results in a file appended with the current date
saveRDS(covid_kulldorff,
  file = here("data", "derived", "public", paste0("covid_kulldorff_", Sys.Date(), ".RDS"))
)
```

Load pre-calculated Kulldorff results

```{r load-Kulldorff}
# load pre-calculated Kulldorff results
# alternatively, modify the file name with an appended date to load a more current set of results
covid_kulldorff <- readRDS(
  here("data", "derived", "public", "covid_kulldorff.RDS")
)
```

Report Kulldorff spatial scan results.

```{r report-Kulldorff}
print("Most likely cluster:", quote = FALSE)
covid_kulldorff$most.likely.cluster
print(
  paste0(
    "Number of Secondary clusters: ",
    length(covid_kulldorff$secondary.clusters)
  ),
  quote = FALSE
)
```

Assign unique cluster ID's to each county within a cluster.
Clusters include the county at the center of a cluster and all of the other counties within the cluster radius.
Therefore, we use the FIPS code of the county at the center of each cluster as the unique cluster ID.

```{r assign-cluster-IDs, message = FALSE}
# list of primary cluster locations (counties)
cluster_locations <- covid_kulldorff$most.likely.cluster$location.IDs.included

# create data frame of clusters and
# calculate the clusterID as the first (center) county FIPS code
clusters <- covid_table[cluster_locations, "fips"] %>%
  mutate(clusterID = covid_table[[cluster_locations[1], "fips"]],
         likelihood = covid_kulldorff$most.likely.cluster$log.likelihood.ratio)

# Get a list of secondary clusters
secondary <- covid_kulldorff$secondary.clusters

# similarly add counties in each secondary cluster to the list of clusters
for (i in secondary) {
  cluster_locations <- i$location.IDs.included
  new_clusters <- covid_table[cluster_locations, "fips"] %>%
    mutate(clusterID = covid_table[[cluster_locations[1], "fips"]],
           likelihood = i$log.likelihood.ratio)
  clusters <- clusters %>% rbind(new_clusters)
}

rm(cluster_locations, secondary, i, new_clusters)
```

### Map: Spatial Clusters

```{r join-clusterID-to-acs_covid}
acs_covid <- acs_covid %>%
  left_join(clusters, by = "fips") %>%
  mutate(isCluster = case_when(
    clusterID == fips ~ "center of cluster",
    !is.na(clusterID) ~ "other part of cluster",
    .default = NA
  ))
```

```{r map-clusters}
tm_spatialepi_clusters <-
  tm_shape(state) +
    tm_fill("gray98") +
  tm_shape(acs_covid) +
  tm_polygons(col = "isCluster",
          palette = "-Oranges",
          popup.vars = c("fips", "clusterID"),
          colorNA = NULL,
          title = "SpatialEpi Kulldorff COVID-19 Clusters",
          border.col = "white",
          lwd = 0.2,
          border.alpha = 0.2) +
  tm_shape(state) +
    tm_borders("grey", lwd = 0.5) +
  tm_layout(
    legend.position = c("left", "bottom"),
    legend.title.size = 0.8,
    legend.text.size = 0.5
  )

tm_spatialepi_clusters
```

### Step 9: Calculate relative risk

`(local cases/local population) / (global cases - local cases/global population - local population)`

Classify counties using `mutate(case_when())`

The `SpatialEpi` implementation of Kulldorff spatial scan statistics does not calculate local relative risk or cluster relative risk.
Therefore, the next step is to calculate local and cluster relative risk (workflow step 7).

```{r relative-risk}
total_pop <- sum(acs_covid$pop)
total_cases <- sum(acs_covid$cases)

acs_covid <- acs_covid %>%
  group_by(clusterID) %>%
  mutate(
    rr_cluster = ifelse(is.na(clusterID), NA,
      (sum(cases) / sum(pop)) / ((total_cases - sum(cases)) / (total_pop - sum(pop)))
    )
  ) %>%
  ungroup() %>%
  mutate(
    rr_loc = (cases / pop) / ((total_cases - cases) / (total_pop - pop))
  )

rm(total_pop, total_cases)
```

Classify relative risk on a scale from 1 to 6

Risk is classified according to this table:

| Relative Risk Values | Relative Risk Class |
|:--------------------:|:-------------------:|
|  Outside of cluster  |          1          |
|       RR \< 1        |          1          |
|    1 \<= RR \< 2     |          2          |
|    2 \<= RR \< 3     |          3          |
|    3 \<= RR \< 4     |          4          |
|    4 \<= RR \< 5     |          5          |
|    5 \<= RR \< 6     |          6          |

Counties falling outside of any cluster are assigned a score of 1.

```{r classify-relative-risk}
# class breaks
breaks <- c(-Inf, 1, 2, 3, 4, 5, Inf)

acs_covid <- acs_covid %>%
  mutate(
    cluster_class = ifelse(is.na(clusterID), 1, cut(rr_cluster, breaks, labels = FALSE)),
    loc_class = cut(rr_loc, breaks, labels = FALSE)
  )
```

### Step 10: Concatenate

`c(state + relative risk)` to produce unique cluster identification codes

### Check number of clusters - compare to total of 102 in original study

```{r make-gee-clusters}
# calculate clusters
acs_covid <- acs_covid %>% mutate(
  rp_clusID = as.integer(statefp) * 10 + cluster_class
)

# summarize clustering results
cluster_summary <- acs_covid %>%
  dplyr::filter(cases > 0) %>%
  st_drop_geometry() %>%
  count(rp_clusID)
cat(
  length(cluster_summary$n),
  "unique clusters based on spatialEpi CLUSTER relative risk\n"
)
summary(cluster_summary$n)
rm(cluster_summary)
```

### Map: Qualitative map of clusters

### Step 11: Compare data with GEE results (`Aug1GEE`)

# Results

Present maps, takeaways from comparing data with original study results.

# Discussion

Explain weaknesses in original study design/data, points for improvement.

# Acknowledgements

This report is based upon the template for Reproducible and Replicable Research in Human-Environment and Geographical Sciences, [DOI:[10.17605/OSF.IO/W29MQ](DOI:%5B10.17605/OSF.IO/W29MQ){.uri}](https://doi.org/10.17605/OSF.IO/W29MQ)

# References
[github.com/HEGSRR/RPr-Chakraborty2021](https://github.com/HEGSRR/RPr-Chakraborty2021)